---
title: Mapa actualizado del coronavirus en el Ecuador
authors: 
  - Alonso Quijano
---

```{r, echo=FALSE, include=FALSE}
# --------- these are the packages used to create the map ---------
library(tidyverse)
library(sf)
library(rmapshaper)
library(tmap)
library(leaflet)
```

```{r, include=FALSE}
# --------- preparing the st data ---------
# --------- Download the shapefile and the coronavirus csv data from github ---------
# The coronavirus data by province and cantons ----

url <- "https://raw.githubusercontent.com/aquijanoruiz/elquantificador_posts/master/salud/2020-03-22-mapa-del-coronavirus-en-el-ecuador/COVID19_ECU_CONFIRMED_CANTON.csv"
confirmed_c <- read_csv(url) # Now we have loaded the confirmed cases by canton

url <- "https://raw.githubusercontent.com/aquijanoruiz/elquantificador_posts/master/salud/2020-03-22-mapa-del-coronavirus-en-el-ecuador/COVID19_ECU_CONFIRMED_PROVINCE.csv"
confirmed_p <- read_csv(url) # Now we have loaded the confirmed cases by province

# Ecuador's administrative division (*.shp) data ----

# Now, we need to load Ecuador's administrative division by province
url <- "https://github.com/aquijanoruiz/elquantificador_posts/raw/master/salud/2020-03-22-mapa-del-coronavirus-en-el-ecuador/ECU_PROVINCE_SHAPEFILE.zip"
td <- tempdir() # We create a temporary directory
tf <- tempfile(tmpdir=td, fileext = ".zip") # We create the placeholder file
download.file(url,tf) # We download the data into the placeholder file

# We get the name of the file inside the zip file that contains the demographic and economic data, 
# unzip it, get the full path name of it, and finally load it
shp.file.name <- unzip(tf, list=TRUE)$Name[4] # The shp file name
shx.file.name <- unzip(tf, list=TRUE)$Name[6] # The shx file name
dbf.fine.name <- unzip(tf, list=TRUE)$Name[8] # The dbf file name
prf.fine.name <- unzip(tf, list=TRUE)$Name[10] # The prf file name

unzip(tf, files=c(shp.file.name, shx.file.name, dbf.fine.name, prf.fine.name), exdir=td, overwrite=TRUE)
shp.file.path <- file.path(td, shp.file.name)

ecu_map_p <- st_read(shp.file.path) # Now we have loaded the shapefile

# We first need to load Ecuador's administrative division by canton
url <- "https://github.com/aquijanoruiz/elquantificador_posts/raw/master/salud/2020-03-22-mapa-del-coronavirus-en-el-ecuador/ECU_CANTON_SHAPEFILE.zip"
td <- tempdir() # We create a temporary directory
tf <- tempfile(tmpdir=td, fileext = ".zip") # We create the placeholder file
download.file(url,tf) # We download the data into the placeholder file

# We get the name of the file inside the zip file that contains the demographic and economic data, 
# unzip it, get the full path name of it, and finally load it
shp.file.name <- unzip(tf, list=TRUE)$Name[9] # The shp file name
shx.file.name <- unzip(tf, list=TRUE)$Name[11] # The shx file name
dbf.fine.name <- unzip(tf, list=TRUE)$Name[5] # The dbf file name
prf.fine.name <- unzip(tf, list=TRUE)$Name[7] # The prf file name

unzip(tf, files=c(shp.file.name, shx.file.name, dbf.fine.name, prf.fine.name), exdir=td, overwrite=TRUE)
shp.file.path <- file.path(td, shp.file.name)

ecu_map_c <- st_read(shp.file.path) # Now we have loaded the shapefile
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, comment=FALSE, include=FALSE}
# --------- preparing canton map ---------
# We simplify the data
ecu_map_c <- ms_simplify(ecu_map_c, keep=0.01) # We keep the 0.4% of the polygon

# We keep only the variables we need
ecu_map_c <- ecu_map_c %>% select(-c(ID_0, ISO, NAME_0, NAME_1, ID_1, ID_2, TYPE_2, ENGTYPE_2, NL_NAME_2, VARNAME_2)) %>%
  rename(Cantón = NAME_2) %>% # We only keep the province names and the geometry
  filter(!Cantón == "Bolívar") # We take out the canton Bolivar beacuse there are two catons with the same name,
  # one in Manabi and one in Carchi. This duplicate creates a problem when illustrating the map. As there are very few
  # cases in Bolivar in Manabi, we will remove them.

# --------- preparing province map ---------
# We simplify the data
ecu_map_p <- ms_simplify(ecu_map_p, keep=0.01) # We keep the 0.4% of the polygon

# We keep only the variables we need
ecu_map_p <- ecu_map_p %>% select(-c(ID_0, ISO, NAME_0, ID_1, TYPE_1, ENGTYPE_1, NL_NAME_1, VARNAME_1)) %>%
  rename(Provincia = NAME_1)

# --------- preparing coronavirus data ---------
# The canton data
confirmed_c

# We need to transform the data from wide to long
confirmed_c <- confirmed_c %>% select(-Provincia) %>% gather(Fecha, Casos, -Cantón) %>% 
  mutate(Fecha = as.Date(Fecha, format = "%m/%d/%y"), Casos = as.numeric(Casos)) %>%
  filter(Fecha >= as.Date("2020-03-16")) %>% # We only have data by province from May 16th
  replace_na(list(Casos = 0)) %>% filter(!Cantón == "Bolívar")

confirmed_c$Cantón <- factor(confirmed_c$Cantón, levels = levels(ecu_map_c$Cantón))

# The province data
confirmed_p

# We need to transform the data from wide to long
confirmed_p <- confirmed_p %>% gather(Fecha, Casos, -Provincia) %>% 
  mutate(Fecha = as.Date(Fecha, format = "%m/%d/%y")) %>%
  filter(Fecha >= as.Date("2020-03-16")) %>% # We only have data by province from May 16th
  replace_na(list(Casos = 0))

confirmed_p$Provincia <- factor(confirmed_p$Provincia, levels = levels(ecu_map_p$Provincia))

# --------- merging the st data with the coronavirus data ---------
centroid <- st_centroid(ecu_map_p)
centroid[23,] # We use Tungurahua province as the center of our map 

# The canton data
covid19_confirmed_c <- inner_join(ecu_map_c, confirmed_c, by = "Cantón") # We merge the map data with the coronavirus data

today <- "2020-04-08"
covid19_confirmed_c <- covid19_confirmed_c %>% filter(Fecha == today) %>% select(-Fecha)

# The province data
covid19_confirmed_p <- inner_join(ecu_map_p, confirmed_p, by = "Provincia") # We merge the map data with the coronavirus data
covid19_confirmed_p <- covid19_confirmed_p %>% filter(Fecha == today) %>% select(-Fecha)


# --------- creating the interactive map ---------
# We filter the data and take out the nonzero elements
covid19_nozeros_c <- covid19_confirmed_c[!covid19_confirmed_c$Casos == 0,]

covid19_confirmed_today_map <-
  tm_shape(ecu_map_p) + tm_borders(col = "grey", lwd = 2, alpha = 0.4) + # This is the province sf file
  tm_polygons(col = "skyblue", alpha = 0.2) + 
  tm_shape(covid19_confirmed_p) +  # We add a second tm_shape with the province data stored in the sf file
  tm_fill(col = "Casos", alpha = 0, legend.show = FALSE) + # We set transparency to 0 because we don't want any colors, just the data
  tm_shape(covid19_nozeros_c) + # We add a third tm_shape with the canton data stored in the sf file
  tm_bubbles(size = "Casos", col = "red", alpha = 0.6, scale = 4, border.lwd = NA) # We add the bubbles
  
covid19_confirmed_today_map <- tmap_leaflet(covid19_confirmed_today_map) # We transform the map into a leaflet map
```

Después que la OMS haya declarado pandemia al brote de Covid-19, el presidente del Ecuador, Lenín Moreno, declaró emergencia sanitaria a todo el país. A fin de luchar contra la propagación del coronavirus, es imporante permanecer en casa, obedeciendo a las medidas de prevención y manteniéndonos informados de cómo avanza la situación sanitaria. Te presentamos un mapa interactivo para que sigas día a día el número de casos confirmados de coronavirus en el país. El mapa está actualizado al **miércoles, 8 de abril de 2020**.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.height = 7, fig.width = 6}
url <- "https://raw.githubusercontent.com/aquijanoruiz/elquantificador_posts/master/salud/2020-03-22-mapa-del-coronavirus-en-el-ecuador/COVID19_ECU_CONFIRMED_NATIONAL.csv"
confirmed_n <- read_csv(url)

# Organizing the data
colnames <- confirmed_n[["Casos"]]
confirmed_n <- t(confirmed_n[,2:ncol(confirmed_n)])
confirmed_n <- cbind(rownames(confirmed_n), confirmed_n)
rownames(confirmed_n) <- NULL
colnames(confirmed_n) <- c("Fecha", colnames)
confirmed_n <- as.data.frame(confirmed_n)
confirmed_n$Fecha <- as.Date(confirmed_n$Fecha, format = "%m/%d/%y")

# We create the table that will illustrate the data
table <- confirmed_n %>% filter(Fecha == today) %>% select(-Fecha)

library(kableExtra)
kable(table) %>% kable_styling() %>% column_spec(1, bold = T) %>%
  column_spec(2, color = "brown", bold = T) %>% column_spec(3, color = "green", bold = T) %>% 
  column_spec(4, color = "orange", bold = T) %>% 
  column_spec(5, color = "skyblue", bold = T)
```

<style>
.html-widget {
    margin: auto;
}
</style>

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.height = 7, fig.width = 6}
covid19_confirmed_today_map %>% removeLayersControl() %>%
  setView(lng = -78.50374, lat = -1.289527, zoom = 6) %>% fitBounds(-80.0, -4.2, -76.6, 0.8) # We center the map in Tungurahua
```

Fuente: Servicio Nacional de Gestión de Riesgo y Emergencia

Este mapa fue elaborado utilizando lenguaje R a través los paquetes `tmap` y `leaflet`. Si estás interesado en reproducir este código y la metodología empleada, visita nuestra página en [github](https://github.com/aquijanoruiz/elquantificador_posts/tree/master/salud/2020-03-22-mapa-del-coronavirus-en-el-ecuador). Para más información, también me puedes escribir a alonso@elquantificador.org.